Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-27 sections/eval.tex
shrink stuff more
Update on Overleaf.

The inner-product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM. Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns, while outer-product computes the product of all pairs of elements from the input matrices.
Gustavson's algorithm computes the output matrix one row at a time, by scaling and merging rows of the second sparse matrix corresponding to the nonzero columns of a row of the first sparse matrix.
Further explanation is available in Section 2.2 of  ~\cite{zhang2021gamma}.

Figures \ref{fig:spgemm_small} and \ref{fig:spgemm_large} illustrate SpGEMM results with Finch and TACO. Notably, inner-product can be very inefficient unless sparse matrices are nearly dense due to the cost of intersections incurring redundant traversals. Additionally, TACO only supports outer product when the output matrix format is dense. It does not support outer product due to its lack of support for scatter into a 2D workspace, limiting its use for large sparse matrices. In contrast, Finch supports scatter into 2D workspace, allowing for comparison with Gustavson's algorithm for large sparse matrices.

As depicted in Figure \ref{fig:spgemm_small}, Finch achieves comparable performance with TACO for inner-product. For large sparse matrices, as shown in Figure \ref{fig:spgemm_large}, Finch outperforms TACO with an average speedup of 1.05. It's worth mentioning that TACO's Gustavson's algorithm is hard-wired whereas Finch's data structure-driving programming enables to systematically generate SpGEMM using explicit temporary, bytemap format, and matrix transposition.

@@

To evaluate the capabilities of Finch in graph applications, we its performance for two fundamantal graph applications: Breadth-first search (BFS) and Bellman-Ford single-source shortest path. We used graphs in GraphBLAS~\cite{mattson2019lagraph}, which include road networks (SNAP/roadNet-CA and GAP/GAP-road) where node degrees are bounded and a diametr is very long and scale-free graphs (the others) where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{beamer2012direction} is crucial to achieve high-performance. Direction-optimized BFS uses push and pull traversal to efficiently explore graphs. Push traversal starts from the source node, pushing neighboring nodes into the queue. Pull traversal starts from the target node, exploring predecessors (parents). This approach uses early-break in pull traversal, which is used to terminate the traversal prematurely once the target node is encountered, to avoid unnecessary edge traversal.

Figure~\cite{graph_result} (a) \changwan{add (a) in fig} shows BFS speedup over Graphs.jl, a BFS implementation with push traversal of Julia. `finch\_push\_only` only uses push traversal, whereas `finch\_push\_pull` applies direction-optimization used in GraphBLAS. Finch supports push/pull traversal, early break, so it supports direction-optimization. Finch additionaly supports the different loop orders and multiple outputs, which also improves the performance. As shown in Figure~\cite{graph_result} (a), direction-optimization significantly improves the performance for scale-free graphs. \changwan{how much average speedup?} GraphBLAS sometimes better than Finch since GraphBLAS incorporates all hardwired optimizations. We believe we will achieve a better performance if we incorporate those optimizations.

\begin{figure}
	\includegraphics[width=\linewidth]{bfs_speedup_over_graphs.jl.png}
	\includegraphics[width=\linewidth]{bellmanford_speedup_over_graphs.jl.png}
    \caption{Performance of graph apps across various tools.}
     \label{graph_result}
\end{figure}

@@

\subsubsection{Evaluating Finch Logic}
To demonstrate the performance of Finch Logic, we evaluate it on a series of kernels which benefit from the kind of kernel fusion that it automatically applies; 1) triangle counting on graphs 2) SDDMM 3) multiple elementwise operations. Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined execution and even takes advantage of vector instructions which are not yet present in Finch.
\begin{figure}
\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
  \vspace{0pt} \includegraphics[width=145pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
  \vspace{0pt} \includegraphics[width=145pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
  \vspace{0pt} \includegraphics[width=145pt, height=120pt]{figures/elementwise_speedup_over_duckdb.png}
\end{tabular}
\caption{Performance of Finch Logic for common kernels.}
\end{figure}

>>>>

The inner-product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM. Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns, while outer-product computes the product of all pairs of elements from the input matrices.
Gustavson's algorithm computes the output matrix one row at a time, by scaling and merging rows of the second sparse matrix corresponding to the nonzero columns of a row of the first sparse matrix.
Further explanation is available in Section 2.2 of~\cite{zhang2021gamma}.

Figures~\ref{fig:spgemm_small} and~\ref{fig:spgemm_large} illustrate SpGEMM results with Finch and TACO. Notably, inner-product can be very inefficient unless sparse matrices are nearly dense due to the cost of intersections incurring redundant traversals. Additionally, TACO only supports outer product when the output matrix format is dense. It does not support outer product due to its lack of support for scatter into a 2D workspace, limiting its use for large sparse matrices. So, we were unable to obtain numbers for inner-product and TACO's outer-product for large matrices. In contrast, Finch supports scatter into 2D workspace, allowing for comparison with Gustavson's algorithm for large sparse matrices.

As depicted in Figure~\ref{fig:spgemm_small}, Finch achieves comparable performance with TACO for inner-product. For large sparse matrices, as shown in Figure~\ref{fig:spgemm_large}, Finch outperforms TACO with an average speedup of 1.05. It's worth mentioning that TACO's Gustavson's algorithm is hard-wired whereas Finch's data structure-driving programming enables to systematically generate SpGEMM using explicit temporary, bytemap format, and matrix transposition.

@@

To evaluate the capabilities of Finch in graph applications, we assessed its performance in two fundamental graph algorithms: Breadth-First Search (BFS) and Bellman-Ford single-source shortest path. The graphs used were obtained from GraphBLAS~\cite{mattson2019lagraph}, including road networks (SNAP/roadNet-CA and GAP/GAP-road), characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short. Direction-optimization~\cite{beamer2012direction} is crucial for achieving high performance in such scenarios.

Direction-optimized BFS utilizes push and pull traversal techniques to efficiently explore graphs. Push traversal originates from the source node, pushing neighboring nodes into the queue, while pull traversal starts from the target node, exploring predecessors (parents). Incorporating early-break in pull traversal enables premature termination upon encountering the target node, thereby avoiding unnecessary edge traversal.

Figure \ref{graph_result} (a) illustrates BFS speedup over Graphs.jl, a BFS implementation with push traversal in Julia. finch\_push\_only exclusively utilizes push traversal, while finch\_push\_pull applies direction-optimization akin to GraphBLAS. Finch's support for push/pull traversal and early break facilitates direction-optimization. Additionally, Finch's support for different loop orders and multiple outputs further enhances performance. Direction-optimization notably enhances performance for scale-free graphs (average speedup to be included). Although GraphBLAS sometimes outperforms Finch due to its incorporation of all hardwired optimizations, we anticipate achieving better performance with Finch upon incorporating similar optimizations. \changwan{mark captions (a) and (b)}

Figure \ref{graph_result} (b) presents Bellman-Ford speedup (with path lengths and shortest-path tree) over Graphs.jl. Among GraphBLAS's five variants for Bellman-Ford, we selected LAGraph\_BF\_full1a, consistently the fastest with our graphs. Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 1.05). Note that we did not include GAP-road as it takes too long time with Graph.jl. \changwan{(self)dataset name changed?}

\begin{figure}
	\includegraphics[width=0.5\linewidth]{bfs_speedup_over_graphs.jl.png}%
	\includegraphics[width=0.5\linewidth]{bellmanford_speedup_over_graphs.jl.png}
    \caption{Performance of graph apps across various tools.}
     \label{graph_result}
\end{figure}

@@

\subsubsection{Evaluating Finch Logic}
To demonstrate the performance of Finch Logic, we evaluate it on a series of kernels which benefit from the kind of kernel fusion that it automatically applies; 1) triangle counting on graphs 2) SDDMM 3) and element-wise operations. Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined query execution. For the element-wise operations, we also provide an unfused Finch implementation and the native Julia implementation for reference
\begin{figure}
\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
  \vspace{0pt} \includegraphics[width=145pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
  \vspace{0pt} \includegraphics[width=145pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
  \vspace{0pt} \includegraphics[width=145pt, height=120pt]{figures/elementwise_speedup_over_duckdb.png}
\end{tabular}
\caption{Performance of Finch Logic for common kernels.}
\end{figure}

### 2024-03-27 sections/lang.tex
Update on Overleaf.

Our syntax is highly permissive: by allowin blocks of code with multiple statements, we support both multiple outputs and temporary tensors, seen most prominently in our implementation of gustavson's algorithm for sparse-sparse matrix multiply and breadth-first-search.
%
Our breadth-first seach also uses an
\mintinline{julia}{if} statement to avoid operating on vertices outside the
frontier. 
%
We have found that \mintinline{julia}{if} is particularly useful in
our model for masking to avoid work.
%
However, to compile these constructs efficient, we must restrict this syntax in several ways, as we have alluded to in the previous section via life cycles.
%
In the semantics section, we detail other specifics that allow us to specialize our syntax efficient code over structured data.

@@

@@

    \begin{prooftree}
    \hypo{\langle val, (e, t) \rangle
    \rightarrow val'}
    \hypo{var\not\in e}
    \infer2[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t) \rangle}}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle args_i, (e, t) \rangle \Rightarrow vals_i}
        \hypo{\langle f, (e, t) \rangle \Rightarrow g}
        \infer2[$Call$]{\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \llangle g(vals...), t \rrangle}
    \end{prooftree}
    \vspace{6pt}

@@

    \begin{prooftree}
        \hypo{\langle head, s_1 \rangle \rightarrow s_2}
        \infer1[$BlockHead$]{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \infer0[$EmptyBlock$]{\langle\finchblock(), s\rangle \rightarrow s}
    \end{prooftree}
    \vspace{6pt}

>>>>

Our syntax is highly permissive: by allowing blocks of code with multiple statements, we support both multiple outputs and temporary tensors, seen most prominently in our implementation of gustavson's algorithm for sparse-sparse matrix multiply and breadth-first-search.
%
Our breadth-first seach also uses an
\mintinline{julia}{if} statement to avoid operating on vertices outside the
frontier. 
%
We have found that \mintinline{julia}{if} is particularly useful in
our model for masking to avoid work.
%
However, to compile these constructs efficient, we must restrict this syntax in several ways, as we have alluded to in the previous section via life cycles.
%
In the semantics section, we detail other specifics that allow us to specialize our syntax efficient code over structured data.

@@

We include the sketch of our semantics to highlight several key design choices.
%
\teo{Make this true:}
First, we note that the life cycle rules (in combination with the definition rules) prevent programs that read and write to values within a loop except in the narrow circumstance of updates, which cannot even be applied to scalar definitions.
%
Beyond the simple management of iterator state, these rules allow us to effectively analyze the propagation of constants within a loop.
%
We eliminate problems that in a less restrictive language would require alias analysis so that we can tell when a zero or repetitive value read from one tensor is being used in an expression or written to another tensor.
%
These restrictions play a similar role to that of a pure function in Halide~\cite{ragan-kelley_halide_2013}.
%
Second, loops with tensor accesses enter into a looplet system via the Unfurl rule.
%
In this system, repeated structures  (Run) and constant informa

@@

% \begin{multline*}
%         Modes:=update \mid read \mid undefined \mid defined
% \end{multline*}
    \begin{prooftree}
    \hypo{\langle val, (e, t) \rangle
    \rightarrow val'}
    \hypo{var\not\in e}
    \infer2[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t) \rangle}}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle args_i, (e, t) \rangle \Rightarrow vals_i}
        \hypo{\langle f, (e, t) \rangle \Rightarrow g}
        \infer2[$Call$]{\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \llangle g(vals...), t \rrangle}
    \end{prooftree}
    \vspace{6pt}

@@

    \begin{prooftree}
    \hypo{\langle head,s_{1} \rangle \rightarrow s_2}
    \infer1[$BlockHead$]{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \infer0[$EmptyBlock$]{\langle\finchblock(), s\rangle \rightarrow s}
    \end{prooftree}
    \vspace{6pt}
